\chapter{Concept and Design}
\label{cha:conceptanddesign}

\section{Language primitives}

I intend to implement an intermediate representation of a Matlab program.
The additional abstraction layer allows language independent optimizations and makes the system independent from the actually used frontend language.
The set of language primitives of the intermediate representation has to include the operational primitives of linear algebra as well as an iteration abstraction in order to realize algorithms based on convergence.
Furthermore, it is of particular interest to keep this set as small as possible, because this would alleviate a possible optimization step prior to execution.

\section{Runtime}

I want to research how the most common linear algebra operations can be efficiently represented within a parallel data flow system.
One key aspect is going to be the matrix matrix multiplication because it is one of the central operations of many machine learning algorithms and it is one of the most time consuming operations as well.
Therefore, it is crucial to efficiently implement this operation.
This requires a thorough investigation and evaluation of all possible implementations.
Since this will strongly depend on the actual distributed system, I will implement and evaluate these operations within Stratosphere.
The Stratosphere implementation acts at the same time as a proof of concept of Gilbert.
I have chosen Stratosphere over other systems, such as Spark, because of its expressiveness and its powerful parallel data-flow optimizer.
Last but not least, Stratosphere is developed at the chair of Prof. Markl, where I write my master's thesis.
However, there should not be any fundamental obstacle which prevent the utilization of an other distributed computing system.
