\chapter{Evaluation}
\label{cha:evaluation}

\chapterquote{An ounce of performance is worth pounds of promises.}{Mae West, (1893 - 1980)}

\begin{itemize}
	\item Evaluate scalability of system for increasing data volume
	\item Evaluate scalability of system for increasing number of nodes
	\item Compare algorithms implemented in Gilbert with specialized version
	\begin{itemize}
		\item Pagerank
		\item K-Means
		\item NNMF
	\end{itemize}
	\item Compare Stratosphere vs. Spark execution engine
	\item Compare Breeze vs. Mahout Math-Backend
\end{itemize}

Pages $\approx$ 8-12

\section{Scalability}

\section{Gilbert Algorithms vs. Specialized Algorithms}

\section{Stratosphere vs. Spark}

\section{Breeze vs. Mahout Math-Backend}

I want to demonstrate the proper functioning and potential of Gilbert by comparing the runtime of different machine learning algorithms implemented in this system with those implemented in a traditional MapReduce framework.
Hopefully, the results of the automatically parallelized algorithms are comparable to the hand-tuned versions.
Furthermore, I like to show the applicability of the proposed system to web-scale analytical processing.
This will require a thorough examination of its scalability behavior.
As benchmarking algorithms I intend to implement the PageRank~\cite{page:1999a}, non-negative matrix factorization~\cite{seung:anips2001a} and k-means~\cite{macqueen:1967a} algorithm.
These algorithms entail to support the basic linear algebra operations as well as an iteration mechanism.
Furthermore, they are examples of widely used algorithms and thus they emphasize the relevance of the developed system.